---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "60%",
  fig.width = 6,
  fig.height = 4.5,
  dpi = 300,
  dev.args = list(bg = "transparent")
)
```

# R package | dp-gaupe-familiegrupper

<!-- badges: start -->

[![R-CMD-check](https://github.com/miljodirektoratet/dp-gaupe-familiegrupper/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/miljodirektoratet/dp-gaupe-familiegrupper/actions/workflows/R-CMD-check.yaml)
[![lint](https://github.com/miljodirektoratet/dp-gaupe-familiegrupper/actions/workflows/lint.yaml/badge.svg)](https://github.com/miljodirektoratet/dp-gaupe-familiegrupper/actions/workflows/lint.yaml)
<!-- [![Codecov test coverage](https://codecov.io/gh/miljodirektoratet/dp-gaupe-familiegrupper/graph/badge.svg)](https://app.codecov.io/gh/miljodirektoratet/dp-gaupe-familiegrupper) -->
<!-- badges: end -->

R package **gaupefam** for clustering lynx observations into family groups using spatio-temporal analysis. Integrated into the **dp-gaupe-familiegrupper** data pipeline for automated processing via Databricks jobs or Docker cron scheduling.

> **Note on Naming**: This repository is named `dp-gaupe-familiegrupper` (the data pipeline), but the R package itself is called `gaupefam`. When installing, use the repository name; when using the package in R code, use the package name.

**Table of Contents:**

- [Project Overview](#project-overview)
  - [Key Features](#key-features)
  - [Repository Structure](#repository-structure)
  - [Docker Configuration](#docker-configuration)
  - [Workflow Statuses](#workflow-statuses)
- [Getting Started](#getting-started)
  - [Pipeline Deployment Options](#pipeline-deployment-options)
  - [Development Installation](#development-installation)
- [Development](#development)
  - [Prerequisites](#prerequisites)
  - [Development Setup](#development-setup)
  - [Development Workflow](#development-workflow)
  - [Version Pinning and Releases](#version-pinning-and-releases)
  - [Code Quality Standards](#code-quality-standards)
  - [GHA Workflows](#gha-workflows)
  - [Branch Protection](#branch-protection)
- [Documentation](#documentation)
- [Acknowledgements](#acknowledgements)
- [License](#license)

## Project Overview

This R package is developed and maintained by the Norwegian Environment Agency (NEA) together with the Norwegian Institute for Nature Research (NINA). It provides clustering functionality for grouping lynx observations into family groups based on spatio-temporal criteria.

**Data Pipeline Integration**: The package is a core component of NEA's **dp-gaupe-familiegrupper** data product pipeline. Feature engineering is performed upstream using Python on Databricks, while this package handles the R-based cluster analysis. The pipeline runs either as:

- **Scheduled Databricks jobs** for automated processing
- **Docker cron jobs** for containerized deployment scenarios

**Workflow**: Python feature engineering (Databricks) → R clustering analysis (this package) → Output to unity catalog.

For more information about the data product, see the [Confluence page](https://miljodir.atlassian.net/wiki/spaces/DDO/pages/11201347593/Rovbase+gaupe+familiegrupper+ikke+i+IT-produktoversikten).

### Key Features

- **Lynx clustering analysis**: Groups lynx observations into family units using spatio-temporal criteria
- **Pipeline integration**: Designed for integration with Databricks and Docker-based data pipelines
- **Reproducible environments**: Docker containers and renv for consistent package versions
- **Quality assurance**: Automated testing, code quality checks, and security scanning via GitHub Actions

### Repository Structure

Important configurations files are listed below, full overview of the repository structure is available in the [Repository Structure](./docs/repo-structure.md) documentation.

| File/Directory            | Purpose                             |
|---------------------------|-------------------------------------|
| `.devcontainer/`          | VS Code dev container configuration |
| `.github/workflows/`      | GitHub Actions for CI/CD            |
| `.pre-commit-config.yaml` | Pre-commit hook configuration       |
| `renv/`                   | renv package management             |
| `renv.lock`               | Locked package versions             |
| `DESCRIPTION`             | Package metadata                    |

### Docker Configuration

This project includes multiple Docker configurations to support different development and deployment scenarios:

- **RStudio Server (dev)**: Web-based development environment running on <http://localhost:8787>
- **VS Code/Codespaces (dev)**: Development environment with Radian console and VS Code extensions running in Codespaces or local VS Code with Dev Containers extension
- **Production (prod)**: Minimal runtime container with only package dependencies

For detailed development container configuration and customization instructions, see the [setup guide](./docs/setup-guide.md#development-container-technical-details).

### Workflow Statuses

| Job | Status | Description |
|---|---|---|
| **R-CMD-check** | ![Status](https://img.shields.io/github/actions/workflow/status/miljodirektoratet/dp-gaupe-familiegrupper/R-CMD-check.yaml?branch=main&label=&style=flat) | R package validation, testing, cross-platform checks |
| **Lint** | ![Status](https://img.shields.io/github/actions/workflow/status/miljodirektoratet/dp-gaupe-familiegrupper/lint.yaml?branch=main&label=&style=flat) | Code style and quality checks using lintr |
| **CI Docker** | ![Status](https://img.shields.io/github/actions/workflow/status/miljodirektoratet/dp-gaupe-familiegrupper/ci-docker-prod.yaml?branch=main&label=&style=flat) | Build and test Docker image |
| **CD Docker** | ![Status](https://img.shields.io/github/actions/workflow/status/miljodirektoratet/dp-gaupe-familiegrupper/cd-docker-prod.yaml?label=&style=flat) | Container deployment to GitHub Registry |
| **Dependabot** | | Automated dependency updates and vulnerability scanning |

## Getting Started

This R package is designed for integration into data pipelines and is installed directly from GitHub. Choose your deployment method based on your pipeline infrastructure:

### Pipeline Deployment Options

#### Databricks Integration (Recommended)

For scheduled data processing in Databricks environments:

```R
# Install in Databricks R notebook ensure to pin to a specific release for stability
install.packages("remotes")
remotes::install_github("miljodirektoratet/dp-gaupe-familiegrupper@v1.0.0")

# Pin to an exact commit SHA
remotes::install_github("miljodirektoratet/dp-gaupe-familiegrupper@<commit_sha>")

# Load and use in your analysis
library(gaupefam)
packageVersion("gaupefam")
# Your clustering analysis code here...
```

**Scheduling**: Configure Databricks jobs to run R notebooks containing the clustering analysis on a scheduled basis (daily, weekly, etc.).

#### Docker Cron Integration

For containerized environments with cron-based scheduling that access Unity Catalog on Databricks (not yet tested):

```bash
# Run pipeline with data stored in Unity Catalog
docker run --rm \
  -e DATABRICKS_HOST="https://your-databricks-instance.cloud.databricks.com" \
  -e DATABRICKS_TOKEN="your-token" \
  ghcr.io/miljodirektoratet/dp-gaupe-familiegrupper:latest \
  Rscript docker-entrypoint-databricks.R

# Run pipeline with data stored in a local volume
docker run --rm \
  -v /path/to/local/data:/data \
  ghcr.io/miljodirektoratet/dp-gaupe-familiegrupper:latest \
  Rscript docker-entrypoint-local-volume.R
```

**Scheduling**: Set up cron jobs to run the Docker container at desired intervals.

```bash
# Schedule with cron (example)
# 0 2 * * * docker run --rm -e DATABRICKS_HOST="$DATABRICKS_HOST" -e DATABRICKS_TOKEN="$DATABRICKS_TOKEN" ghcr.io/miljodirektoratet/dp-gaupe-familiegrupper:latest Rscript docker-entrypoint.R
```

### Development Installation

For package development or testing outside of pipeline environments install the package using `devtools`:

```{r example, fig.width = 5, fig.height = 3.5, dpi = 300, out.width = "50%"}
# Initialize renv (if not already done)
renv::init()

# Check status with dev packages loaded
renv::status(dev = TRUE)

# Load package
devtools::load_all()

# Load and test basic functionality
library(gaupefam)
gaupefam::hello()
#> 🚀 Hello from gaupefam !

# Test plotting capabilities
map <- gaupefam::plot_trondelag()
map
```

## Development
<!-- For contributors -->

> [!TIP]
> **Recommended Setup:** Use [VS Code Devcontainers](https://code.visualstudio.com/docs/devcontainers/containers) for a consistent containerized environment. Alternative options available are:
>
> - **RStudio Devcontainer**: Web-based RStudio Server in Docker
> - **Local Setup**: Native environment using `renv` package management
>
> For RStudio devcontainer or local setup instructions, see the comprehensive [setup guide](./docs/setup-guide.md).

The following steps configure your development environment using VS Code Dev Containers.

### Prerequisites
<!-- Software versions, system requirements, dependencies, account requirements (APIs, services) -->

- [Docker](https://docs.docker.com/engine/install/)
- [Visual Studio Code](https://code.visualstudio.com/)
- [VS Code Dev Containers](https://code.visualstudio.com/docs/devcontainers/containers)
extension

### Development Setup
<!-- Step-by-step installation guide -->

1. **Open the project**: Clone and open the project folder in VS Code or use [GitHub Codespaces](https://docs.github.com/en/codespaces).

2. **Start the devcontainer**: When prompted, reopen the folder in the Devcontainer. If not prompted, manually trigger it via the Command Palette (`Ctrl+Shift+P` or `Cmd+Shift+P`) and select "Dev Containers: Reopen in Container".

3. **Automatic setup**: The devcontainer automatically:
   - Configures VS Code with recommended settings and extensions per [devcontainer.json](.devcontainer/devcontainer.json)
   - Installs development tools: [Git](https://git-scm.com/), [pre-commit](https://pre-commit.com/)
   - Sets up the R environment with dependencies in `renv` via `renv::restore()`
   - Executes initial code quality checks (`devtools::check()`) and tests (`devtools::test()`)

4. **Verify the installation**:

   ```r
   # Load the package
   devtools::load_all()

   # Run tests
   devtools::test()

   # Run full package checks
   devtools::check()

   # Run complete local CI workflow
   source("scripts/ci-local.R")
   ```

5. **Test the example notebook**: Open `notebooks/example_family_grouping_workflow.Rmd` in RStudio or render it with `rmarkdown::render("notebooks/example_family_grouping_workflow.Rmd")` to see a complete workflow demonstration.

6. **Configure the GitHub Repository**: If you fork this repository or use it to create a new repository from scratch, you'll need to ensure your GitHub repository is correctly configured. This repository has no secrets and variables configured for GitHub Actions, but you should verify that your **security scans** are properly set up. See the **GitHub Repository Configuration** section in the [setup guide](./docs/setup-guide.md) for instructions.

### Development Workflow

1. **Create branch**: create a new `feat/<name>` or bug fix `fix/<name>` branch using Git.
2. **Write Code**: develop your package functions in `R/`.
3. **Document**: use `roxygen2` comments and run `devtools::document()`.
4. **Test**: write tests in `tests/testthat/` and run `devtools::test()`.
5. **Check**: run `devtools::check()` to ensure your package passes all checks.
6. **Local CI**: run the complete quality check workflow with `source("scripts/ci-local.R")` which executes:
   - `renv::status()` and `renv::snapshot()` for dependency management
   - `devtools::load_all()` to test package loading
   - `lintr::lint_dir()` for code quality
   - `styler::style_dir()` for code formatting
   - `devtools::document()` and `devtools::build_readme()` for documentation
   - `devtools::test()` and `devtools::check()` for validation
7. **README**: update `README.Rmd` and build with `devtools::build_readme()`.
8. **Commit**: stage and commit changes using Git, pre-commit hooks will run automatically.
9. **Push**: push your branch to GitHub and create a pull request for review, GitHub Actions will run tests automatically.

See [Command Cheat Sheet](./docs/command-cheatsheet.md) for common commands and `scripts/ci-local.R` for the complete local CI workflow.

### Release Workflow

Versioning is defined in `DESCRIPTION` (`Version:`).
Development versions typically use the `.9000` suffix (for example `0.0.0.9000`), while releases use semantic versions (`X.Y.Z`) and a matching git tag (`vX.Y.Z`).

Release helper script:

```bash
# From repository root
chmod +x scripts/release_workflow.sh
./scripts/release_workflow.sh
```

The script will:

1. Check that you are on `main` with a clean git working tree.
2. Prompt for a release version (for example `1.0.0`).
3. Update `DESCRIPTION` to that version and commit it.
4. Create and push tag `v1.0.0` (which triggers Docker CD with semver tags).
5. Bump `DESCRIPTION` to the next development version (`1.1.0.9000`) and push.

Recommended first release flow:

1. Merge your final release-ready changes into `main`.
2. Pull latest `main` locally.
3. Run `./scripts/release_workflow.sh` and enter `1.0.0`.
4. Verify GitHub Actions runs in the Actions tab.
5. Install and verify the pinned release:

```r
remotes::install_github(
  "miljodirektoratet/dp-gaupe-familiegrupper@v1.0.0",
  upgrade = "never"
)
packageVersion("gaupefam")
```

### Code Quality Standards

Follow R coding standards with roxygen2 documentation, consistent formatting via styler, and comprehensive testing with testthat. All quality checks are automated through the `scripts/ci-local.R` script, pre-commit hooks, and CI workflows.

See the [Code Quality and Security Standards](./docs/code-and-security-standards.md) guide to see which rules are enforced in this repo.

### GHA Workflows

The repository includes automated workflows for code quality, security, and deployment:

| Workflow | Trigger | Purpose |
|----------|---------|---------|
| **R-CMD-check** | `workflow_dispatch` (manual) | R package validation, testing, cross-platform checks |
| **lint** | `workflow_dispatch` (manual) | Code style and quality checks using lintr |
| **CI Docker** | `push` (all branches except `main`), `pull_request`, `workflow_dispatch` | Docker build and test |
| **CD Docker** | `push` to `main`, semver tags `v*.*.*`, `workflow_dispatch` | Container deployment to GitHub Registry |
| **Dependabot** | `schedule` | Automated dependency updates and vulnerability scanning |

The R package workflows ensure code quality and compatibility across different R versions and operating systems. The **R-CMD-check** workflow is the standard for R package validation, while **lint** ensures consistent code style following the project's configuration.

### Branch Protection

The `main` branch is protected with the following rules:

- Require pull request reviews before merging
- Require status checks to pass before merging
- Require branches to be up to date before merging

## Documentation

- **Data Product Documentation**: [Confluence page](https://miljodir.atlassian.net/wiki/spaces/DDO/pages/11201347593/Rovbase+gaupe+familiegrupper+ikke+i-IT-produktoversikten)
- **Development Guides**: See [`/docs`](./docs/) folder:
  - [Setup Guide](./docs/setup-guide.md)
  - [Code Quality and Security Standards](./docs/code-and-security-standards.md)
  - [Command Cheatsheet](./docs/command-cheatsheet.md)
  - [Repository Structure](./docs/repo-structure.md)

## Acknowledgements

This repository leverages the following tools from the R community:

- **Pre-commit hooks**:
  - [pre-commit-hooks](https://github.com/pre-commit/pre-commit-hooks)
  - [lorenzwalthert/precommit](https://github.com/lorenzwalthert/precommit)
- **Development practices** from [R Packages (2e)](https://r-pkgs.org/)
  by Hadley Wickham and Jenny Bryan
- **Docker images** from the [Rocker
  Project](https://rocker-project.org/)
- **ggmap**: D. Kahle and H. Wickham. ggmap: Spatial Visualization with
  ggplot2. The R Journal, 5(1), 144-161. URL
  <http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf>

## License

The package **dp-gaupe-familiegrupper** is released under the MIT License. See [LICENSE](LICENSE) file for details.

---
