---
title: "Delta Sharing Native R Package"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This notebook demonstrates how to use the native R `delta.sharing` package from GitHub (zacdav-db/delta-sharing-r). This package provides a pure R implementation for Delta Sharing without requiring Python integration.

## Installation

```{r install-package, eval=FALSE}
# Install the native R delta.sharing package from GitHub
# Note: This may require increasing C stack limits inside Docker containers.
# Recommended to run locally and not inside a container if possible.

# Method 1: Standard installation
remotes::install_github("zacdav-db/delta-sharing-r")
```

## Check Installation

```{r check-installation}
# Check if the package is installed and load it
if (requireNamespace("delta.sharing", quietly = TRUE)) {
  library(delta.sharing)
  cat("✅ delta.sharing package loaded successfully\n")
  cat("Package version:", packageVersion("delta.sharing"), "\n")
} else {
  cat("❌ delta.sharing package not installed\n")
  cat("Please run the installation code above\n")
  knitr::knit_exit()
}
```

## Load Additional Libraries

```{r load-libraries}
library(dplyr) # For data manipulation
library(ggplot2) # For plotting
library(jsonlite) # For config file reading
```

## Load Configuration

```{r load-config}
# Read the Delta Sharing configuration
cat("Current working directory:", getwd(), "\n")

config_path <- "../config.share"

# Check if file exists and read it
if (!file.exists(config_path)) {
  # Try absolute path if relative doesn't work
  config_path <- "/home/rstudio/workspace/config.share"
  cat("Trying absolute path:", config_path, "\n")
}

cat("Using config file:", config_path, "\n")
cat("File exists:", file.exists(config_path), "\n")

# Read configuration
config <- jsonlite::fromJSON(config_path)
cat("Endpoint:", config$endpoint, "\n")
cat("Token expires:", config$expirationTime, "\n")

# Check if token is expiring soon
expiry_date <- as.POSIXct(config$expirationTime, format = "%Y-%m-%dT%H:%M:%OSZ", tz = "UTC")
current_date <- Sys.time()
time_until_expiry <- difftime(expiry_date, current_date, units = "hours")

cat("Time until token expires:", round(as.numeric(time_until_expiry), 2), "hours\n")

if (time_until_expiry < 24) {
  cat("⚠️  WARNING: Token expires within 24 hours!\n")
}
```

## Create Delta Sharing Client

```{r create-client}
# Create a Delta Sharing client using the configuration file
# The delta.sharing package should provide a client function

# Check available functions in the package
cat("Available functions in delta.sharing:\n")
print(ls("package:delta.sharing"))

# Create client (syntax will depend on the package implementation)
# Common patterns might be:
tryCatch(
  {
    # Try different possible function names
    if (exists("delta_sharing_client")) {
      client <- delta_sharing_client(config_path)
      cat("✅ Client created with delta_sharing_client()\n")
    } else if (exists("DeltaSharingClient")) {
      client <- DeltaSharingClient(config_path)
      cat("✅ Client created with DeltaSharingClient()\n")
    } else if (exists("sharing_client")) {
      client <- sharing_client(config_path)
      cat("✅ Client created with sharing_client()\n")
    } else {
      cat("❓ Please check package documentation for client creation\n")
      cat("Available functions:\n")
      print(ls("package:delta.sharing"))
    }
  },
  error = function(e) {
    cat("❌ Error creating client:", e$message, "\n")
    cat("Please check the package documentation\n")
  }
)
```

## List Available Shares

```{r list-shares}
# List shares using the native R package
if (exists("client")) {
  tryCatch(
    {
      # Try different possible function names
      if (exists("list_shares")) {
        shares <- list_shares(client)
      } else if (exists("client$list_shares")) {
        shares <- client$list_shares()
      } else {
        cat("❓ Please check documentation for list_shares function\n")
        shares <- NULL
      }

      if (!is.null(shares)) {
        cat("Available shares:\n")
        print(shares)

        # Store first share for later use
        if (nrow(shares) > 0) {
          share_name <- shares$name[1]
          cat("Using share:", share_name, "\n")
        }
      }
    },
    error = function(e) {
      cat("❌ Error listing shares:", e$message, "\n")
    }
  )
} else {
  cat("❌ No client available - please check client creation above\n")
}
```

## List Schemas

```{r list-schemas}
# List schemas in the first share
if (exists("share_name")) {
  tryCatch(
    {
      if (exists("list_schemas")) {
        schemas <- list_schemas(client, share_name)
      } else if (exists("client$list_schemas")) {
        schemas <- client$list_schemas(share_name)
      } else {
        cat("❓ Please check documentation for list_schemas function\n")
        schemas <- NULL
      }

      if (!is.null(schemas)) {
        cat("Available schemas:\n")
        print(schemas)

        if (nrow(schemas) > 0) {
          schema_name <- schemas$name[1]
          cat("Using schema:", schema_name, "\n")
        }
      }
    },
    error = function(e) {
      cat("❌ Error listing schemas:", e$message, "\n")
    }
  )
}
```

## List Tables

```{r list-tables}
# List tables in the first schema
if (exists("schema_name")) {
  tryCatch(
    {
      if (exists("list_tables")) {
        tables <- list_tables(client, share_name, schema_name)
      } else if (exists("client$list_tables")) {
        tables <- client$list_tables(share_name, schema_name)
      } else {
        cat("❓ Please check documentation for list_tables function\n")
        tables <- NULL
      }

      if (!is.null(tables)) {
        cat("Available tables:\n")
        print(tables)

        if (nrow(tables) > 0) {
          table_name <- tables$name[1]
          cat("Using table:", table_name, "\n")
        }
      }
    },
    error = function(e) {
      cat("❌ Error listing tables:", e$message, "\n")
    }
  )
}
```

## Load Table Data

```{r load-data}
# Load actual data from the Delta Share table
if (exists("table_name")) {
  tryCatch(
    {
      # Try different possible function names for data loading
      if (exists("load_table")) {
        df <- load_table(client, share_name, schema_name, table_name)
      } else if (exists("read_table")) {
        df <- read_table(client, share_name, schema_name, table_name)
      } else if (exists("client$load_table")) {
        df <- client$load_table(share_name, schema_name, table_name)
      } else if (exists("delta_sharing_read")) {
        df <- delta_sharing_read(client, share_name, schema_name, table_name)
      } else {
        cat("❓ Please check documentation for data loading function\n")
        df <- NULL
      }

      if (!is.null(df)) {
        cat("✅ Data loaded successfully!\n")
        cat("Data dimensions:", nrow(df), "rows x", ncol(df), "columns\n")
        cat("Column names:", paste(names(df), collapse = ", "), "\n")

        # Show data structure
        cat("\nData structure:\n")
        str(df)

        # Show first few rows
        cat("\nFirst 6 rows:\n")
        print(head(df))
      }
    },
    error = function(e) {
      cat("❌ Error loading data:", e$message, "\n")
    }
  )
}
```

## Data Analysis Example

```{r data-analysis}
# Perform basic data analysis if data was loaded successfully
if (exists("df") && !is.null(df) && nrow(df) > 0) {
  cat("=== DATA ANALYSIS ===\n")

  # Basic summary
  cat("Dataset Summary:\n")
  print(summary(df))

  # Check for missing values
  missing_counts <- sapply(df, function(x) sum(is.na(x)))
  if (any(missing_counts > 0)) {
    cat("\nMissing values per column:\n")
    print(missing_counts[missing_counts > 0])
  }

  # Create visualizations for numeric columns
  numeric_cols <- sapply(df, is.numeric)

  if (any(numeric_cols)) {
    numeric_col <- names(df)[which(numeric_cols)[1]]
    cat("\nCreating histogram for:", numeric_col, "\n")

    p1 <- ggplot(df, aes_string(x = numeric_col)) +
      geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
      labs(title = paste("Distribution of", numeric_col)) +
      theme_minimal()

    print(p1)
  }

  # Show categorical data summary
  char_cols <- sapply(df, function(x) is.character(x) || is.factor(x))
  if (any(char_cols)) {
    char_col <- names(df)[which(char_cols)[1]]
    cat("\nTop values in", char_col, ":\n")
    print(head(sort(table(df[[char_col]]), decreasing = TRUE)))
  }
} else {
  cat("No data available for analysis\n")
}
```

## Package Documentation

```{r package-docs}
# Display package documentation and help
cat("=== PACKAGE INFORMATION ===\n")

if (requireNamespace("delta.sharing", quietly = TRUE)) {
  # Package description
  desc <- packageDescription("delta.sharing")
  cat("Package:", desc$Package, "\n")
  cat("Version:", desc$Version, "\n")
  cat("Title:", desc$Title, "\n")
  cat("Description:", desc$Description, "\n")
  cat("Author:", desc$Author, "\n")

  # List all exported functions
  cat("\nExported functions:\n")
  exported <- getNamespaceExports("delta.sharing")
  for (func in exported) {
    cat("-", func, "\n")
  }

  cat("\nFor detailed help, use:\n")
  cat("help(package = 'delta.sharing')\n")
  cat("?function_name  # for specific function help\n")
}
```

## Alternative: Manual Function Discovery

```{r function-discovery}
# If the package functions aren't obvious, explore them manually
if (requireNamespace("delta.sharing", quietly = TRUE)) {
  cat("=== FUNCTION DISCOVERY ===\n")

  # Get all objects in the package namespace
  all_objects <- ls("package:delta.sharing")

  cat("All objects in package:\n")
  for (obj in all_objects) {
    obj_class <- class(get(obj, envir = asNamespace("delta.sharing")))
    cat("-", obj, "(", paste(obj_class, collapse = ", "), ")\n")
  }

  # Look for common patterns
  client_functions <- all_objects[grepl("client|Client", all_objects)]
  list_functions <- all_objects[grepl("list|List", all_objects)]
  load_functions <- all_objects[grepl("load|read|Load|Read", all_objects)]

  if (length(client_functions) > 0) {
    cat("\nPossible client functions:\n")
    print(client_functions)
  }

  if (length(list_functions) > 0) {
    cat("\nPossible listing functions:\n")
    print(list_functions)
  }

  if (length(load_functions) > 0) {
    cat("\nPossible data loading functions:\n")
    print(load_functions)
  }
}
```

## Troubleshooting Installation

```{r installation-troubleshooting}
cat("=== INSTALLATION TROUBLESHOOTING ===\n\n")

cat("If you encounter issues installing delta.sharing:\n\n")

cat("1. **C Stack Limit Errors**:\n")
cat("   - Increase stack limits before installation\n")
cat("   - Install dependencies separately first\n")
cat("   - Use binary packages where available\n\n")

cat("2. **Arrow Installation Issues**:\n")
cat("   - Set LIBARROW_MINIMAL environment variable\n")
cat("   - Install system dependencies if needed\n")
cat("   - Consider using pre-compiled arrow binaries\n\n")

cat("3. **Alternative Approaches**:\n")
cat("   - Use the Python integration notebook instead\n")
cat("   - Use the REST API notebook for basic functionality\n")
cat("   - Clone and build from source manually\n\n")

cat("4. **renv Integration**:\n")
cat("   After successful installation, run:\n")
cat("   renv::snapshot()  # to capture the package in your lockfile\n\n")

cat("5. **Check Installation**:\n")
cat("   Run: library(delta.sharing) to verify installation\n")
```

## Summary

```{r summary}
cat("=== SUMMARY ===\n\n")

if (requireNamespace("delta.sharing", quietly = TRUE)) {
  cat("✅ Native R delta.sharing package is available\n")
  cat("✅ Provides pure R implementation (no Python required)\n")
  cat("✅ Should integrate well with renv\n")

  if (exists("df") && !is.null(df)) {
    cat("✅ Successfully loaded data:", nrow(df), "rows\n")
  }

  cat("\n🔧 Advantages:\n")
  cat("- Pure R implementation\n")
  cat("- No Python dependencies\n")
  cat("- Better renv integration\n")
  cat("- Native R data types\n")

  cat("\n⚠️ Considerations:\n")
  cat("- GitHub package (not CRAN)\n")
  cat("- May have installation challenges in some environments\n")
  cat("- Smaller community than Python implementation\n")
} else {
  cat("❌ Native R delta.sharing package not installed\n")
  cat("📋 Alternative options:\n")
  cat("1. Use delta_sharing_python_connect.Rmd\n")
  cat("2. Use delta_sharing_rest_api_connect.Rmd\n")
  cat("3. Resolve installation issues and try again\n")
}

cat("\n📚 Resources:\n")
cat("- GitHub: https://github.com/zacdav-db/delta-sharing-r\n")
cat("- Delta Sharing Spec: https://github.com/delta-io/delta-sharing\n")
```

## Notes

### About this Package:
- **Source**: Community-maintained R implementation by zacdav-db
- **Status**: Experimental/development package
- **Dependencies**: arrow, bit, bit64, and other data processing packages

### Installation with renv:
```r
# After successful installation, capture in lockfile:
renv::snapshot()

# In future sessions or other environments:
renv::restore()
```

### For Production Use:
- Test thoroughly in your specific environment
- Consider pinning to a specific commit/version
- Have fallback options (Python or REST API approaches)
- Monitor the GitHub repository for updates

This notebook provides a framework for using the native R delta.sharing package. The exact function names and usage patterns may vary depending on the package version and API design.
