---
title: "Lynx Family Grouping - Complete Workflow"
author: "Miljodirektoratet"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Overview

This notebook demonstrates the complete workflow for grouping lynx observations
into family groups using the `gaupefam` package. The workflow follows a clean
pipeline approach:

1. **Load and explore data** - Understand the spatial and temporal distribution
2. **Group observations** - Use `group_lynx_families()` to assign family groups
3. **Assess sensitivity** - Use `compare_grouping_methods()` to test robustness
4. **Create visualizations** - Generate maps and summaries
5. **Export results** - Save to desired format (GeoPackage, CSV, Unity Catalog, etc.)

## Key Features

- **Pure functions** - No side effects, no hidden file I/O
- **Node pipeline** - Clear sequence of transformations
- **Flexible I/O** - You control where and how to save results
- **Reproducible** - Same inputs always produce same outputs

---

# 1. Setup

## 1.1 Load Libraries

```{r libraries}
# Load required packages
library(gaupefam) # Our package with all family grouping functions
library(sf) # Spatial data handling
library(dplyr) # Data manipulation
library(ggplot2) # Plotting
library(leaflet) # Interactive maps
library(htmlwidgets) # Save interactive maps
library(viridis) # Color palettes

# Enable s2 for spherical geometry
sf::sf_use_s2(TRUE)

# Print software versions
cat("GEOS version:", sf::sf_extSoftVersion()["GEOS"], "\n")
cat("GDAL version:", sf::sf_extSoftVersion()["GDAL"], "\n")
cat("PROJ version:", sf::sf_extSoftVersion()["PROJ"], "\n")
```

## 1.2 Load Test Data

```{r load-data}
# Load test data from package
data(lynx_family_test_data)

# Assign to shorter variable name
dat <- lynx_family_test_data

# Inspect data structure
cat("Number of observations:", nrow(dat), "\n")
print(dat)
```

---

# 2. Explore Data

## 2.1 Spatial Distribution

```{r explore-spatial}
# Plot observation locations
ggplot(dat) +
  geom_sf(aes(color = byttedyr), size = 4) +
  geom_sf_text(aes(label = rovbase_id), vjust = -1, size = 5) +
  scale_color_viridis_d(option = "plasma") +
  theme_minimal() +
  labs(
    title = "Spatial Distribution of Lynx Observations",
    subtitle = paste("n =", nrow(dat), "observations"),
    color = "Prey Class",
    x = "Longitude (SWEREF99 TM)",
    y = "Latitude (SWEREF99 TM)"
  )

# Calculate spatial statistics
coords <- st_coordinates(dat)
cat("\nSpatial extent:\n")
cat("  X range:", round(min(coords[, 1])), "to", round(max(coords[, 1])), "m\n")
cat("  Y range:", round(min(coords[, 2])), "to", round(max(coords[, 2])), "m\n")

# Calculate pairwise distances
dist_matrix <- st_distance(dat)
dist_numeric <- as.numeric(dist_matrix) # Convert units to numeric (meters)
cat("\nPairwise distances:\n")
cat("  Min:", round(min(dist_numeric[dist_numeric > 0])), "m\n")
cat("  Max:", round(max(dist_numeric)), "m\n")
cat("  Mean:", round(mean(dist_numeric[dist_numeric > 0])), "m\n")
```

## 2.2 Temporal Distribution

```{r explore-temporal}
# Create timeline plot
dat_timeline <- dat %>%
  st_drop_geometry() %>%
  mutate(
    obs_id = factor(rovbase_id),
    duration_days = as.numeric(difftime(datotid_til, datotid_fra, units = "days"))
  ) %>%
  arrange(datotid_fra)

ggplot(dat_timeline, aes(y = obs_id)) +
  geom_segment(
    aes(x = datotid_fra, xend = datotid_til, yend = obs_id),
    linewidth = 3, color = "steelblue"
  ) +
  geom_point(aes(x = datotid_fra), size = 4, color = "darkblue") +
  geom_point(aes(x = datotid_til), size = 4, color = "darkred") +
  theme_minimal() +
  labs(
    title = "Temporal Distribution of Lynx Observations",
    subtitle = "Blue = start, Red = end",
    x = "Date",
    y = "Observation ID"
  )

# Temporal statistics
cat("\nTemporal extent:\n")
cat("  First observation:", format(min(dat$datotid_fra), "%Y-%m-%d %H:%M"), "\n")
cat("  Last observation:", format(max(dat$datotid_til), "%Y-%m-%d %H:%M"), "\n")
cat("  Time span:", round(difftime(max(dat$datotid_til), min(dat$datotid_fra), units = "days"), 1), "days\n")
cat("\nObservation durations:\n")
cat("  Min:", round(min(dat_timeline$duration_days), 1), "days\n")
cat("  Max:", round(max(dat_timeline$duration_days), 1), "days\n")
cat("  Mean:", round(mean(dat_timeline$duration_days), 1), "days\n")
```

---

# 3. Group Observations

## 3.1 Basic Grouping

```{r basic-grouping}
# Run grouping with default parameters
grouped_basic <- group_lynx_families(
  data = dat,
  clustering_method = "cluster_custom",
  ordering_method = "time",
  reversed = FALSE,
  optimize_group_count = TRUE,
  optimize_distances = TRUE
)

# Inspect results
cat("Grouping results:\n")
cat("  Total groups:", length(unique(grouped_basic$group_id)), "\n")
cat("  Observations:", nrow(grouped_basic), "\n\n")

# Group size distribution
group_sizes <- table(grouped_basic$group_id)
cat("Group sizes:\n")
print(group_sizes)
cat("\n  Mean group size:", round(mean(group_sizes), 2), "\n")
cat("  Max group size:", max(group_sizes), "\n")
```

## 3.2 Visualize Groups

```{r visualize-groups}
# Plot grouped observations
ggplot(grouped_basic) +
  geom_sf(aes(color = factor(group_id)), size = 5) +
  geom_sf_text(aes(label = rovbase_id), vjust = -1.5, size = 4) +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal() +
  labs(
    title = "Lynx Family Groups",
    subtitle = paste(length(unique(grouped_basic$group_id)), "groups identified"),
    color = "Group ID"
  )
```

## 3.3 Alternative Configurations

```{r alternative-configs}
# Try hierarchical clustering
grouped_hierarchical <- group_lynx_families(
  data = dat,
  clustering_method = "cluster_hierarchical",
  ordering_method = "time",
  optimize_group_count = TRUE,
  optimize_distances = TRUE,
  hclust_poly = 1
)

cat("Hierarchical clustering:\n")
cat("  Total groups:", length(unique(grouped_hierarchical$group_id)), "\n\n")

# Try different ordering
grouped_pca <- group_lynx_families(
  data = dat,
  clustering_method = "cluster_custom",
  ordering_method = "pca1",
  optimize_group_count = TRUE,
  optimize_distances = TRUE
)

cat("PCA1 ordering:\n")
cat("  Total groups:", length(unique(grouped_pca$group_id)), "\n\n")

# Without optimization
grouped_no_opt <- group_lynx_families(
  data = dat,
  clustering_method = "cluster_custom",
  ordering_method = "time",
  optimize_group_count = FALSE,
  optimize_distances = FALSE
)

cat("No optimization:\n")
cat("  Total groups:", length(unique(grouped_no_opt$group_id)), "\n")
```

---

# 4. Sensitivity Analysis

## 4.1 Compare All Methods

```{r compare-methods}
# Run comprehensive comparison
# This tests 30 different combinations:
# - 2 clustering methods (hierarchical, custom)
# - 15 ordering strategies (5 methods × forward/reverse + 5 random)

cat("Running sensitivity analysis (30 configurations)...\n")
cat("This may take a minute...\n\n")

comparison <- compare_grouping_methods(
  data = dat,
  optimize_group_count = TRUE,
  optimize_distances = TRUE
)

# Display results
print(comparison)
```

## 4.2 Analyze Results

```{r analyze-comparison}
# Summary statistics
cat("\nHierarchical clustering:\n")
cat("  Min groups:", min(comparison$n_groups_hierarchical), "\n")
cat("  Max groups:", max(comparison$n_groups_hierarchical), "\n")
cat("  Mean groups:", round(mean(comparison$n_groups_hierarchical), 2), "\n")
cat("  SD:", round(sd(comparison$n_groups_hierarchical), 2), "\n\n")

cat("Custom clustering:\n")
cat("  Min groups:", min(comparison$n_groups_custom), "\n")
cat("  Max groups:", max(comparison$n_groups_custom), "\n")
cat("  Mean groups:", round(mean(comparison$n_groups_custom), 2), "\n")
cat("  SD:", round(sd(comparison$n_groups_custom), 2), "\n\n")

# Best configurations
best_hierarchical <- comparison[which.min(comparison$n_groups_hierarchical), ]
cat("Best hierarchical configuration:\n")
cat("  Method:", best_hierarchical$ordering_method, "\n")
cat("  Reversed:", best_hierarchical$reversed, "\n")
cat("  Groups:", best_hierarchical$n_groups_hierarchical, "\n\n")

best_custom <- comparison[which.min(comparison$n_groups_custom), ]
cat("Best custom configuration:\n")
cat("  Method:", best_custom$ordering_method, "\n")
cat("  Reversed:", best_custom$reversed, "\n")
cat("  Groups:", best_custom$n_groups_custom, "\n")
```

## 4.3 Visualize Sensitivity

```{r plot-sensitivity}
# Reshape for plotting
comparison_long <- comparison %>%
  tidyr::pivot_longer(
    cols = c(n_groups_hierarchical, n_groups_custom),
    names_to = "clustering_method",
    values_to = "n_groups"
  ) %>%
  mutate(
    clustering_method = gsub("n_groups_", "", clustering_method),
    ordering_label = paste0(ordering_method, ifelse(reversed, "_rev", ""))
  )

# Plot comparison
ggplot(comparison_long, aes(x = ordering_label, y = n_groups, fill = clustering_method)) +
  geom_col(position = "dodge") +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Sensitivity Analysis: Number of Groups by Configuration",
    subtitle = paste("Based on", nrow(dat), "observations"),
    x = "Ordering Method",
    y = "Number of Groups",
    fill = "Clustering Method"
  )
```

---

# 5. Create Visualization Objects

## 5.1 Calculate Group Centers

```{r group-centers}
# Use the best configuration
grouped_final <- grouped_basic

# Calculate centroids for each group
group_centers <- create_center_points(
  data = grouped_final,
  group_col = "group_id"
)

cat("Group centers calculated:\n")
print(group_centers)
```

## 5.2 Create Connection Lines

```{r group-lines}
# Create lines connecting observations to group centers
group_lines <- create_lines(
  data = grouped_final,
  centers = group_centers,
  group_col = "group_id"
)

# Add group_id back to lines for visualization
# (create_lines only returns rovbase_id by default)
group_lines$group_id <- grouped_final$group_id[match(group_lines$rovbase_id, grouped_final$rovbase_id)]

cat("Connection lines created:\n")
cat("  Total lines:", nrow(group_lines), "\n")
print(head(group_lines))
```

## 5.3 Interactive Map

```{r interactive-map}
# Transform to WGS84 for leaflet
grouped_wgs84 <- st_transform(grouped_final, 4326)
centers_wgs84 <- st_transform(group_centers, 4326)
lines_wgs84 <- st_transform(group_lines, 4326)

# Create color palette
n_groups <- length(unique(grouped_wgs84$group_id))
pal <- colorFactor(viridis(n_groups, option = "turbo"), domain = grouped_wgs84$group_id)

# Create interactive map
lynx_map <- leaflet() %>%
  addProviderTiles(providers$OpenTopoMap) %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  # Add connection lines
  addPolylines(
    data = lines_wgs84,
    color = ~ pal(group_id),
    weight = 2,
    opacity = 0.6,
    group = "Connections"
  ) %>%
  # Add observation points
  addCircleMarkers(
    data = grouped_wgs84,
    color = ~ pal(group_id),
    fillColor = ~ pal(group_id),
    fillOpacity = 0.8,
    radius = 6,
    stroke = TRUE,
    weight = 2,
    popup = ~ paste0(
      "<strong>Observation ID:</strong> ", rovbase_id, "<br>",
      "<strong>Group:</strong> ", group_id, "<br>",
      "<strong>Prey:</strong> ", byttedyr, "<br>",
      "<strong>Date:</strong> ", format(datotid_fra, "%Y-%m-%d")
    ),
    group = "Observations"
  ) %>%
  # Add group centers
  addCircleMarkers(
    data = centers_wgs84,
    color = "black",
    fillColor = ~ pal(group_id),
    fillOpacity = 1,
    radius = 10,
    stroke = TRUE,
    weight = 3,
    popup = ~ paste0("<strong>Group Center</strong><br>Group ID: ", group_id),
    group = "Centers"
  ) %>%
  # Add labels
  addLabelOnlyMarkers(
    data = grouped_wgs84,
    label = ~ as.character(rovbase_id),
    labelOptions = labelOptions(
      noHide = FALSE,
      direction = "top",
      textOnly = TRUE,
      style = list("font-size" = "14px", "font-weight" = "bold")
    ),
    group = "Labels"
  ) %>%
  # Add controls
  addLayersControl(
    baseGroups = c("OpenTopoMap", "Satellite"),
    overlayGroups = c("Observations", "Centers", "Connections", "Labels"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  addMeasure(primaryLengthUnit = "meters", primaryAreaUnit = "sqmeters")

# Display map
lynx_map
```

---

# 6. Export Results

## 6.1 Save Spatial Data

```{r save-spatial}
# Create output directory if it doesn't exist
output_dir <- "output"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Save to GeoPackage (recommended for spatial data)
st_write(
  grouped_final,
  file.path(output_dir, "family_group_observations.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

st_write(
  group_centers,
  file.path(output_dir, "family_group_centers.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

st_write(
  group_lines,
  file.path(output_dir, "family_group_lines.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

cat("Spatial data saved to GeoPackage:\n")
cat("  -", file.path(output_dir, "family_group_observations.gpkg"), "\n")
cat("  -", file.path(output_dir, "family_group_centers.gpkg"), "\n")
cat("  -", file.path(output_dir, "family_group_lines.gpkg"), "\n")
```

## 6.2 Save Tabular Data

```{r save-tabular}
# Save as CSV (without geometry)
results_table <- grouped_final %>%
  st_drop_geometry() %>%
  select(rovbase_id, datotid_fra, datotid_til, byttedyr, group_id)

write.csv(
  results_table,
  file.path(output_dir, "family_group_assignments.csv"),
  row.names = FALSE
)

cat("\nTabular data saved to CSV:\n")
cat("  -", file.path(output_dir, "family_group_assignments.csv"), "\n")
```

## 6.3 Save Comparison Results

```{r save-comparison}
# Save sensitivity analysis results
write.csv(
  comparison,
  file.path(output_dir, "sensitivity_analysis.csv"),
  row.names = FALSE
)

cat("\nSensitivity analysis saved:\n")
cat("  -", file.path(output_dir, "sensitivity_analysis.csv"), "\n")
```

## 6.4 Save Interactive Map

```{r save-map}
# Save interactive map as HTML
saveWidget(
  lynx_map,
  file = file.path(output_dir, "family_groups_map.html"),
  selfcontained = TRUE
)

cat("\nInteractive map saved:\n")
cat("  -", file.path(output_dir, "family_groups_map.html"), "\n")
```

---

# 7. Summary

## 7.1 Results Summary

```{r summary}
cat("===== ANALYSIS SUMMARY =====\n\n")
cat("Dataset:\n")
cat("  Observations:", nrow(dat), "\n")
cat("  Time span:", round(difftime(max(dat$datotid_til), min(dat$datotid_fra), units = "days"), 1), "days\n")
cat("  Spatial extent:", round(max(st_distance(dat))), "m\n\n")

cat("Grouping Configuration:\n")
cat("  Clustering method: cluster_custom\n")
cat("  Ordering method: time\n")
cat("  Optimization: group count + distances\n\n")

cat("Results:\n")
cat("  Family groups identified:", length(unique(grouped_final$group_id)), "\n")
cat("  Observations per group (mean):", round(mean(table(grouped_final$group_id)), 2), "\n")
cat("  Largest group size:", max(table(grouped_final$group_id)), "\n\n")

cat("Sensitivity Analysis:\n")
cat("  Configurations tested: 30\n")
cat("  Groups range (hierarchical):", min(comparison$n_groups_hierarchical), "-", max(comparison$n_groups_hierarchical), "\n")
cat("  Groups range (custom):", min(comparison$n_groups_custom), "-", max(comparison$n_groups_custom), "\n\n")

cat("Output Files:\n")
cat("  - family_group_observations.gpkg\n")
cat("  - family_group_centers.gpkg\n")
cat("  - family_group_lines.gpkg\n")
cat("  - family_group_assignments.csv\n")
cat("  - sensitivity_analysis.csv\n")
cat("  - family_groups_map.html\n\n")

cat("===========================\n")
```

---

# 8. Next Steps

This workflow can be adapted for:

- **Production data**: Replace `lynx_family_test_data` with your own dataset
- **Unity Catalog**: Save results to Delta Lake instead of GeoPackage
- **Custom parameters**: Adjust clustering methods, optimization flags, etc.
- **Batch processing**: Loop over multiple years/regions
- **Quality control**: Add validation checks and diagnostic plots

For more information, see the package documentation:
```r
?group_lynx_families
?compare_grouping_methods
```
