---
title: "Lynx Family Groups Pipeline"
author: "Miljodirektoratet"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

# Lynx Family Groups Analysis Pipeline

This notebook walks through the spatial analysis and grouping of lynx family
observations, using custom clustering and interactive mapping. The main steps include:

1. Loading and exploring test data
2. Comparing grouping methods (sensitivity analysis)
3. Grouping observations with optimal configuration
4. Creating visualization objects
5. Interactive mapping
6. Exporting results (user controls where/how)

## 1. Setup and Load Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# Clear workspace
rm(list = ls())
```

```{r load-libraries}
# Detect base path first (needed for loading package)
if (grepl("/home/rstudio/workspace", getwd())) {
  base_path <- "/home/rstudio/workspace"
} else {
  base_path <- "/home/wilaca/git/miljodirektoratet/dp-gaupe-familiegrupper"
}

cat("Base path set to:", base_path, "\n")
cat("Current working directory:", getwd(), "\n")

# Load gaupefam package (development version)
devtools::load_all(base_path)

# Load other required packages
library(sf) # Spatial data handling
library(dplyr) # Data manipulation
library(leaflet) # Interactive maps
library(htmlwidgets) # Save interactive maps
library(viridis) # Color palettes

# Enable s2 for spherical geometry
sf::sf_use_s2(TRUE)

# Print software versions
cat("GEOS version:", sf::sf_extSoftVersion()["GEOS"], "\n")
cat("GDAL version:", sf::sf_extSoftVersion()["GDAL"], "\n")
cat("PROJ version:", sf::sf_extSoftVersion()["PROJ"], "\n")
```

## 2. Load and Explore Data

```{r load-data, fig.width=10, fig.height=8}
# Load test data from gaupefam package
data(lynx_family_test_data)

# Assign to shorter variable name
dat <- lynx_family_test_data

# Check the data
cat("Number of observations:", nrow(dat), "\n")
print(dat)

# Simple plot of test data
par(mfrow = c(1, 2))

# Plot 1: Spatial distribution
plot(st_geometry(dat),
  pch = 19, col = "red", cex = 2,
  main = "Spatial Distribution", axes = TRUE
)
text(st_coordinates(dat),
  labels = dat$rovbase_id,
  pos = 3, cex = 1
)

# Plot 2: Temporal distribution
plot(dat$datotid_fra, 1:nrow(dat),
  pch = 19, col = "blue", cex = 2,
  main = "Temporal Distribution",
  xlab = "Date", ylab = "Observation ID",
  yaxt = "n"
)
axis(2, at = 1:nrow(dat), labels = dat$rovbase_id)
segments(dat$datotid_fra, 1:nrow(dat),
  dat$datotid_til, 1:nrow(dat),
  col = "blue", lwd = 2
)

par(mfrow = c(1, 1))
```

## 3. Compare Grouping Methods (Sensitivity Analysis)

```{r compare-methods}
# Test 30 different configurations to find optimal grouping:
# - 2 clustering methods (hierarchical, custom)
# - 15 ordering strategies (time, pca1, pca2, north-south, east-west × forward/reverse + 5 random)

cat("Running sensitivity analysis (30 configurations)...\n\n")

comparison <- compare_grouping_methods(
  data = dat,
  optimize_group_count = TRUE,
  optimize_distances = TRUE
)

# Display results
print(comparison)

# Summary statistics
cat("\nHierarchical clustering:\n")
cat("  Min groups:", min(comparison$n_groups_hierarchical), "\n")
cat("  Max groups:", max(comparison$n_groups_hierarchical), "\n")
cat("  Mean groups:", round(mean(comparison$n_groups_hierarchical), 2), "\n\n")

cat("Custom clustering:\n")
cat("  Min groups:", min(comparison$n_groups_custom), "\n")
cat("  Max groups:", max(comparison$n_groups_custom), "\n")
cat("  Mean groups:", round(mean(comparison$n_groups_custom), 2), "\n\n")

# Find best configurations
best_hierarchical <- comparison[which.min(comparison$n_groups_hierarchical), ]
cat("Best hierarchical configuration:\n")
cat("  Method:", best_hierarchical$ordering_method, "\n")
cat("  Reversed:", best_hierarchical$reversed, "\n")
cat("  Groups:", best_hierarchical$n_groups_hierarchical, "\n\n")

best_custom <- comparison[which.min(comparison$n_groups_custom), ]
cat("Best custom configuration:\n")
cat("  Method:", best_custom$ordering_method, "\n")
cat("  Reversed:", best_custom$reversed, "\n")
cat("  Groups:", best_custom$n_groups_custom, "\n")
```

## 4. Group Observations with Best Configuration

```{r group-observations}
# Determine best overall configuration
if (best_hierarchical$n_groups_hierarchical < best_custom$n_groups_custom) {
  best_method <- "cluster_hierarchical"
  best_ordering <- best_hierarchical$ordering_method
  best_reversed <- best_hierarchical$reversed
  n_groups <- best_hierarchical$n_groups_hierarchical
} else {
  best_method <- "cluster_custom"
  best_ordering <- best_custom$ordering_method
  best_reversed <- best_custom$reversed
  n_groups <- best_custom$n_groups_custom
}

cat("Using best configuration:\n")
cat("  Clustering method:", best_method, "\n")
cat("  Ordering method:", best_ordering, "\n")
cat("  Reversed:", best_reversed, "\n")
cat("  Expected groups:", n_groups, "\n\n")

# Run grouping with optimal configuration
grouped_lynx <- group_lynx_families(
  data = dat,
  clustering_method = best_method,
  ordering_method = best_ordering,
  reversed = best_reversed,
  optimize_group_count = TRUE,
  optimize_distances = TRUE,
  group_col = "gruppe_id" # Norwegian naming convention
)

# Display results
cat("Grouping results:\n")
cat("  Total groups:", length(unique(grouped_lynx$gruppe_id)), "\n")
cat("  Total observations:", nrow(grouped_lynx), "\n\n")

# Group size distribution (detailed)
group_sizes <- table(grouped_lynx$gruppe_id)
cat("Group sizes (observations per group):\n")
for (i in seq_along(group_sizes)) {
  cat("  Group", names(group_sizes)[i], ":", group_sizes[i], "observations\n")
}
cat("\n")
```

## 5. Create Visualization Objects

```{r create-viz-objects}
# Extract group assignments (rovbase_id: gruppe_id)
group_assignments <- grouped_lynx %>%
  st_drop_geometry() %>%
  select(rovbase_id, gruppe_id)

# Join group assignments to "original" observations dataset for visualization
lynx_family_observations <- lynx_family_test_data %>%
  left_join(group_assignments, by = "rovbase_id")

# Calculate group centers (centroids) based on clustered observations
lynx_family_group_centers <- create_center_points(
  data = grouped_lynx,
  group_col = "gruppe_id"
)

cat("Group centers calculated:\n")
print(lynx_family_group_centers)

# Create lines connecting observations to group centers
lynx_family_group_lines <- create_lines(
  observations = lynx_family_observations, # Observations with gruppe_id
  centers = lynx_family_group_centers, # Group centroids
  group_col = "gruppe_id"
)

# Add gruppe_id back to lines for visualization
lynx_family_group_lines$gruppe_id <- lynx_family_observations$gruppe_id[match(lynx_family_group_lines$rovbase_id, lynx_family_observations$rovbase_id)]
cat("\nConnection lines created:\n")
cat("  Total lines:", nrow(lynx_family_group_lines), "\n")
```

## 6. Interactive Map Visualization

```{r visualize-map, fig.width=10, fig.height=8}
# Transform to WGS84 for leaflet
grouped_wgs84 <- st_transform(lynx_family_observations, 4326)
centers_wgs84 <- st_transform(lynx_family_group_centers, 4326)
lines_wgs84 <- st_transform(lynx_family_group_lines, 4326)

# Create color palette
n_groups_final <- length(unique(grouped_wgs84$gruppe_id))
pal <- colorFactor(viridis(n_groups_final, option = "turbo"), domain = grouped_wgs84$gruppe_id)

# Create interactive map
lynx_map <- leaflet() %>%
  addProviderTiles(providers$OpenTopoMap) %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  # Add connection lines
  addPolylines(
    data = lines_wgs84,
    color = ~ pal(gruppe_id),
    weight = 3,
    opacity = 0.8
  ) %>%
  # Add observation points
  addCircleMarkers(
    data = grouped_wgs84,
    color = "black",
    fillColor = ~ pal(gruppe_id),
    fillOpacity = 0.8,
    radius = 6,
    stroke = TRUE,
    weight = 2,
    popup = ~ paste0(
      "<strong>Observation ID:</strong> ", rovbase_id, "<br>",
      "<strong>Group:</strong> ", gruppe_id, "<br>",
      "<strong>Prey:</strong> ", byttedyr, "<br>",
      "<strong>Date:</strong> ", format(datotid_fra, "%Y-%m-%d")
    ),
    label = ~ as.character(rovbase_id)
  ) %>%
  # Add group centers
  addCircleMarkers(
    data = centers_wgs84,
    color = "red",
    fillColor = "red",
    fillOpacity = 1,
    radius = 8,
    stroke = TRUE,
    weight = 2,
    popup = ~ paste0("<strong>Group Center</strong><br>Group ID: ", gruppe_id),
    label = ~ paste("Group", gruppe_id)
  ) %>%
  # Add controls
  addLayersControl(
    baseGroups = c("OpenTopoMap", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  addMeasure(primaryLengthUnit = "meters", primaryAreaUnit = "sqmeters")

# Display map
lynx_map
```

## 7. Export Results

```{r export-results}
# Create output directory if it doesn't exist
output_dir <- file.path(base_path, "output")
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Save to GeoPackage (recommended for spatial data)
st_write(
  grouped_lynx,
  file.path(output_dir, "lynx_family_group_observations_migrated.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

st_write(
  lynx_family_group_centers,
  file.path(output_dir, "lynx_family_group_centers_migrated.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

st_write(
  lynx_family_group_lines,
  file.path(output_dir, "lynx_family_group_lines_migrated.gpkg"),
  delete_dsn = TRUE,
  quiet = TRUE
)

cat("Spatial data saved to GeoPackage:\n")
cat("  -", file.path(output_dir, "lynx_family_group_observations_migrated.gpkg"), "\n")
cat("  -", file.path(output_dir, "lynx_family_group_centers_migrated.gpkg"), "\n")
cat("  -", file.path(output_dir, "lynx_family_group_lines_migrated.gpkg"), "\n\n")

# Save as CSV (without geometry)
results_table <- grouped_lynx %>%
  st_drop_geometry() %>%
  select(rovbase_id, datotid_fra, datotid_til, byttedyr, gruppe_id)

write.csv(
  results_table,
  file.path(output_dir, "lynx_family_group_assignments_migrated.csv"),
  row.names = FALSE
)

cat("Tabular data saved to CSV:\n")
cat("  -", file.path(output_dir, "lynx_family_group_assignments_migrated.csv"), "\n\n")

# Save comparison results
write.csv(
  comparison,
  file.path(output_dir, "sensitivity_analysis_migrated.csv"),
  row.names = FALSE
)

cat("Sensitivity analysis saved:\n")
cat("  -", file.path(output_dir, "sensitivity_analysis_migrated.csv"), "\n\n")

# Save interactive map
saveWidget(
  lynx_map,
  file = file.path(output_dir, "lynx_family_groups_map_migrated.html"),
  selfcontained = TRUE
)

cat("Interactive map saved:\n")
cat("  -", file.path(output_dir, "lynx_family_groups_map_migrated.html"), "\n")
```
